---
title: "ISLR Chapter 4 - applied"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR)
library(MASS)
library(class)
library(ggplot2)
library(dplyr)
library(tidyr)
```


## Exercise 10

**This question should be answered using the `Weekly` data set, which is part of the `ISLR` package. It contains 1,089 weekly returns for 21 years, from the beginning of 1990 to the end of 2010.**

**Produce some numerical and graphical summaries of the `Weekly` data. Do there appear to be any patterns?**
```{r}
lags <- Weekly %>% 
  select(-Volume, -Today) %>% 
  gather(key = Lag, value = Returns, -Direction, -Year) %>% 
  ggplot() +
  geom_boxplot(aes(x = Lag, y = Returns, fill = Direction)) +
  theme_minimal()

vol_returns <- ggplot(data = Weekly, aes(x = Volume, y = Today)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

vol_year <- Weekly %>% 
  group_by(Year) %>% 
  summarise(Volume = sum(Volume)) %>% 
  ggplot() +
  geom_line(aes(x = Year, y = Volume)) +
  theme_minimal()

returns <- mutate(Weekly, Week = 1:n()) %>% 
  ggplot(aes(x = Week, y = Today)) +
  geom_line() +
  theme_minimal()

gridExtra::grid.arrange(lags, vol_returns, vol_year, returns, nrow = 2, ncol = 2)
```

Volumes have gone up considerably since 1990, but levelling off in about 2002 to 2003, and a drop from 2008. The box plots show that returns appear to fluctuate around zero; where stocks have gone down this week, returns were more frequently higher last week and five weeks ago, than stocks that went up. The opposite is true for two weeks, with higher returns for stocks that went up this week, and there is no clear relationship for 3 and 4 week lags. There is no relationship between trading volumes and current returns. Returns are more volatile in the recent part of the data set, with significantly negative returns past the 900th observation. 

**Use the full data set to perform a logistic regression with `Direction` as the response and the five lag variables plus `Volume` as predictors. Use the summary function to print the results. Do any of the predictors appear to be statistically significant? If so, which ones?**
```{r}
fit <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
           family = "binomial",
           data = Weekly)

summary(fit)
```

Only `Lag2` appears to be significant at the 95% level. The positive coefficient suggests that positive returns 2 weeks ago are associated with positive returns today. 

**Compute the confusion matrix and overall fraction of correct predictions. Explain what the confusion matrix is telling you about the types of mistakes made by logistic regression.**
```{r}
probs <- predict(fit, type = "response")
preds <- rep("Down", nrow(Weekly))
preds[probs > 0.5] <- "Up"
table(preds, Weekly$Direction)
mean(preds == Weekly$Direction)
```

The predictions appear to suggest that the logistic regression model predicts a high number of false positives - i.e. when the market goes down, logistic regression usually predicts up. The opposite problem is not the case here - the model rarely predicts the market going down where the actual result was up. The overall fraction of correct predictions suggests that the model is slightly better than a coin flip.

**Now fit the logistic regression model using a training data period from 1990 to 2008, with `Lag2` as the only predictor. Compute the confusion matrix and the overall fraction of correct predictions for the held out data (that is, the data from 2009 and 2010).**
```{r}
train <- subset(Weekly, Weekly$Year <= 2008)
fit2 <- glm(Direction ~ Lag2, data = train, family = "binomial")
summary(fit2)
test <- subset(Weekly, Weekly$Year > 2008)
probs2 <- predict(fit2, test, type = "response")
preds2 <- rep("Down", nrow(test))
preds2[probs2 > 0.5] <- "Up"
table(preds2, test$Direction)
mean(test$Direction == preds2)
```

**Repeat (d) using LDA.**
```{r}
fit3 <- lda(Direction ~ Lag2, data = train)
fit3
preds3 <- predict(fit3, test)
table(preds3$class, test$Direction)
mean(preds3$class == test$Direction)
```

As expected LDA results are similar to logistic regression. 

**Repeat (d) using QDA.**
```{r}
fit4 <- qda(Direction ~ Lag2, data = train)
fit4
preds4 <- predict(fit4, test)
table(preds4$class, test$Direction)
mean(preds4$class == test$Direction)
```

QDA results are not good - it is outperformed by LDA and logistic regression and always picks 'Up'

**Repeat (d) using KNN with $K$ = 1.**
```{r}
train_direction <- train[["Direction"]]
train_mat <- as.matrix(train[["Lag2"]])
test_mat <- as.matrix(test[["Lag2"]])
preds5 <- knn(train_mat, test_mat, train_direction, k = 1)
table(preds5, test[["Direction"]])
mean(preds5 == test$Direction)
```

KNN with $K$ = 1 perfoms worst of all, it's essentially a coin flip. 

**Which of these methods appears to provide the best results on this data?**

LDA and logistic regression performed best in predicting the test observations.

**Experiment with different combinations of predictors, including possible transformations and interactions, for each of the methods. Report the variables, method, and associated confu- sion matrix that appears to provide the best results on the held out data. Note that you should also experiment with values for $K$ in the KNN classifier.**

Try interaction terms with `Lag1` and `Volume`:
```{r}
glm.fit.interaction <- glm(Direction ~ Lag1*Lag2, data = train, family = "binomial")
prob.interaction <- predict(glm.fit.interaction, test, type = "response")
pred.interaction <- ifelse(prob.interaction > 0.5, "Up", "Down")
table(pred.interaction, test$Direction)
mean(pred.interaction == test$Direction)
```
Including an interaction term with logistic regression gives us a lower accuracy than the single predictor model.

Now try with an additional predictor, `Lag1`:
```{r}
glm.fit.lag1 <- glm(Direction ~ Lag1 + Lag2, data = train, family = "binomial")
prob.lag1 <- predict(glm.fit.lag1, test, type = "response")
pred.lag1 <- ifelse(prob.lag1 > 0.5, "Up", "Down")
table(pred.lag1, test$Direction)
mean(pred.lag1 == test$Direction)
```

Again, including `Lag1` also worsens performance.

Now try transforming `Lag2`:
```{r}
glm.fit.sq <- glm(Direction ~ poly(Lag2, 2), data = train, family = "binomial")
prob.sq <- predict(glm.fit.sq, test, type = "response")
pred.sq <- ifelse(prob.sq > 0.5, "Up", "Down")
table(pred.sq, test$Direction)
mean(pred.sq == test$Direction)
```

A square transformation appears to perform just as well as the single predictor model. 

Now try these variations with LDA and QDA models:
```{r}
lda.inter <- lda(Direction ~ Lag1*Lag2, data = train)
pred.inter <- predict(lda.inter, test)
table(pred.inter$class, test$Direction)
mean(pred.inter$class == test$Direction)

lda.lag1 <- lda(Direction ~ Lag1 + Lag2, data = train)
pred.lag1 <- predict(lda.lag1, test)
table(pred.lag1$class, test$Direction)
mean(pred.lag1$class == test$Direction)

lda.sq <- lda(Direction ~ poly(Lag2, 2), data = train)
pred.sq <- predict(lda.sq, test)
table(pred.sq$class, test$Direction)
mean(pred.sq$class == test$Direction)
```
The LDA models also don't outperform the single variable models, but of these, that including the square transformation performs best. 

Now try QDA:
```{r}
qda.inter <- qda(Direction ~ Lag1*Lag2, data = train)
pred.inter <- predict(qda.inter, test)
table(pred.inter$class, test$Direction)
mean(pred.inter$class == test$Direction)

qda.lag1 <- qda(Direction ~ Lag1 + Lag2, data = train)
pred.lag1 <- predict(qda.lag1, test)
table(pred.lag1$class, test$Direction)
mean(pred.lag1$class == test$Direction)

qda.sq <- qda(Direction ~ poly(Lag2, 2), data = train)
pred.sq <- predict(qda.sq, test)
table(pred.sq$class, test$Direction)
mean(pred.sq$class == test$Direction)
```

Of the three models, the square transformation performs best, equalling the single predictor LDA and logistic regression test accuracy. 

Now try a few different values for $k$ in a KNN classifier:
```{r}
set.seed(1)
k5pred <- knn(train_mat, test_mat, train_direction, k = 5)
table(k5pred, test$Direction)
mean(k5pred == test$Direction)

k10pred <- knn(train_mat, test_mat, train_direction, k = 10)
table(k10pred, test$Direction)
mean(k10pred == test$Direction)

k20pred <- knn(train_mat, test_mat, train_direction, k = 20)
table(k20pred, test$Direction)
mean(k20pred == test$Direction)
```
None of the KNN classifiers perform particularly well - choosing `Up` every time would perform better than any of them. They are clearly outperformed by the single predictor LDA and logistic regression classifiers. 

## Exercise 11
**In this problem, you will develop a model to predict whether a given car gets high or low gas mileage based on the `Auto` data set.**

**Create a binary variable, `mpg01`, that contains a 1 if `mpg` contains a value above its median, and a 0 if `mpg` contains a value below its median. You can compute the median using the `median()` function. Note you may find it helpful to use the `data.frame()` function to create a single data set containing both `mpg01` and the other Auto variables.**
```{r}
mpg01 <- ifelse(Auto$mpg > median(Auto$mpg), 1, 0)
auto_new <- data.frame(Auto, mpg01)
```

**Explore the data graphically in order to investigate the association between `mpg01` and the other features. Which of the other features seem most likely to be useful in predicting `mpg01`? Scatterplots and boxplots may be useful tools to answer this question. Describe your findings.**
```{r}
disp <- ggplot(data = auto_new) +
  geom_boxplot(aes(x = as.factor(mpg01), y = displacement)) +
  theme_minimal()

horse <- ggplot(data = auto_new) +
  geom_boxplot(aes(x = as.factor(mpg01), y = horsepower)) +
  theme_minimal()

weight <- ggplot(data = auto_new) +
  geom_boxplot(aes(x = as.factor(mpg01), y = weight)) +
  theme_minimal()

acc <- ggplot(data = auto_new) +
  geom_boxplot(aes(x = as.factor(mpg01), y = acceleration)) +
  theme_minimal()

gridExtra::grid.arrange(disp, horse, weight, acc, ncol = 2, nrow = 2)
```

More fuel efficient cars are on average lighter, have lower displacement and horsepower, and better acceleration than less fuel efficient cars. 

```{r}
cyl <- ggplot(data = auto_new) +
  geom_point(aes(x = cylinders, y = mpg)) +
  theme_minimal()

year <- ggplot(data = auto_new) +
  geom_point(aes(x = year, y = mpg)) +
  theme_minimal()

origin <- ggplot(data = auto_new) +
  geom_point(aes(x = origin, y = mpg)) +
  theme_minimal()

gridExtra::grid.arrange(cyl, year, origin, ncol = 3, nrow = 1)
```

`cylinders` is negatively associated with `mpg`, new cars are associated with higher `mpg`, and Japanese cars are most fuel efficient, followed by European, then American. 

**Split the data into a training set and a test set.**
```{r}
set.seed(1)
auto_new$origin <- as.factor(auto_new$origin)
test_index <- sample(nrow(auto_new), 0.2 * nrow(auto_new))
test <- auto_new[test_index,]
train <- auto_new[-test_index,]
```

**Perform LDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?**
```{r}
lda.fit <- lda(mpg01 ~ cylinders + displacement + horsepower + weight + acceleration + year + origin, data = auto_new)
lda.fit
lda.pred <- predict(lda.fit, test) 
mean(lda.pred$class != test$mpg01)
```
LDA gives a test error rate of 6.4%.

**Perform QDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?**
```{r}
qda.fit <- qda(mpg01 ~ cylinders + displacement + horsepower + weight + acceleration + year + origin, data = auto_new)
qda.fit
qda.pred <- predict(qda.fit, test)
mean(qda.pred$class != test$mpg01)
```
QDA gives a slightly higher test error of 7.7%.

**Perform logistic regression on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?**
```{r}
glm.fit <- glm(mpg01 ~ cylinders + displacement + horsepower + weight + acceleration + year + origin, data = auto_new, family = "binomial")
summary(glm.fit)
glm.pred <- predict(glm.fit, test, type = "response")
preds <- rep(0, nrow(test))
preds[glm.pred > 0.5] <- 1
mean(preds != test$mpg01)
```

Logistic regression also gives a test error rate of about 7.7%.

**Perform KNN on the training data, with several values of K, in order to predict `mpg01`. Use only the variables that seemed most associated with `mpg01` in (b). What test errors do you obtain? Which value of $K$ seems to perform the best on this data set?**

Trying KNN for different values of $K$:
```{r}
auto_new$us <- ifelse(auto_new$origin == 1, 1, 0)
auto_new$eu <- ifelse(auto_new$origin == 2, 1, 0)
train <- auto_new[-test_index,]
test <- auto_new[test_index,]

train_mat <- as.matrix(train[,-which(names(train) %in% c("name", "mpg01", "origin"))])
test_mat <- as.matrix(test[,-which(names(test) %in% c("name", "mpg01", "origin"))])

train_mpg01 <- train[, "mpg01"]

set.seed(1)
test_errs <- sapply(1:314, function(i) {
  knn.preds <- knn(train_mat, test_mat, train_mpg01, k = i)
  mean(knn.preds != test$mpg01)
})

which(test_errs == min(test_errs))
```

$K$ = 3 gives the lowest test error rate of `r min(test_errs)` but, judging from the plot below, some much higher values of $K$ still perform quite well. We may prefer to choose one of these values as a classifier with a higher $K$ is likely to have lower variance.

```{r}
plot(1:314, test_errs, type = "l", xlab = "K", ylab = "Test error")
```

## Exercise 12

**This problem involves writing functions.**

**Write a function, `Power()`, that prints out the result of raising 2 to the 3rd power. In other words, your function should compute 2^3^ and print out the results.**
_Hint: Recall that `x^a` raises `x` to the power `a`. Use the `print()` function to output the result._

```{r}
Power <- function() {
  print(2^3)
}

Power()
```

**Create a new function, `Power2()`, that allows you to pass any two numbers, `x` and `a`, and prints out the value of `x^a`. You can do this by beginning your function with the line**
```
Power2=function(x,a){
```
**You should be able to call your function by entering, for instance,**
```
Power2(3 ,8)
```
**on the command line. This should output the value of 3^8^, namely, 6,561.**

```{r}
Power2 <- function(x, a) {
  print(x^a)
}

Power2(3, 8)
```

**Using the `Power2()` function that you just wrote, compute 10^3^, 8^17^, and 131^3^.**
```{r}
Power2(10, 3)
Power2(8, 17)
Power2(131, 3)
```

**Now create a new function, `Power3()`, that actually returns the result `x^a` as an `R` object, rather than simply printing it to the screen. That is, if you store the value `x^a` in an object called `result` within your function, then you can simply `return()` this result, using the following line:**
```
return(result)
```
**The line above should be the last line in your function, before the `}` symbol.**
```{r}
Power3 <- function(x, a) {
  result <- x^a
  return(result)
}
```

**Now using the `Power3()` function, create a plot of $f(x)=x^2$. The x-axis should display a range of integers from 1 to 10, and the y-axis should display $x^2$. Label the axes appropriately, and use an appropriate title for the figure. Consider displaying either the x-axis, the y-axis, or both on the log-scale. You can do this by using `log = ''x''`, log=‘‘y’’, or `log=''xy''` as arguments to the `plot()` function.**
```{r}
plot(log(1:10), log(Power3(1:10, 2)), main = "Square function plot", xlab = "X", ylab = "X^2", xaxt= "n")
axis(side = 1, at = log(1:10), labels = 1:10)
```

**Create a function, `PlotPower()`, that allows you to create a plot of `x` against `x^a` for a fixed `a` and for a range of values of `x`. For instance, if you call**
```
PlotPower (1:10 ,3)
```
**then a plot should be created with an x-axis taking on values 1,2,...,10, and a y-axis taking on values 1^3^,2^3^,...,10^3^.**

```{r}
PlotPower <- function(x, a) {
  plot(x, Power3(x, a), main = "Power function plot", xlab = "X", ylab = paste0("X^", a), xaxt= "n")
  axis(side = 1, at = x)
}
PlotPower(1:10, 3)
```

## Exercise 13
**Using the `Boston` data set, fit classification models in order to predict whether a given suburb has a crime rate above or below the median. Explore logistic regression, LDA, and KNN models using various sub-sets of the predictors. Describe your findings.**

Construct the response variable from `medv`:
```{r}
set.seed(1)
Boston$hiCrime <- factor(ifelse(Boston$crim > median(Boston$crim), "High", "Low"), levels = c("Low", "High"))
```

Summary statistics for `Boston`:
```{r}
summary(Boston)
```

Split the data into a training and test set:
```{r}
train_idx <- sample(nrow(Boston), 0.7*nrow(Boston))
train <- Boston[train_idx,]
test <- Boston[-train_idx,]
```

First try a logistic regression with all predictors:
```{r}
glm.fit <- glm(hiCrime ~ . - crim, data = train, family = "binomial")
summary(glm.fit)
glm.prob <- predict(glm.fit, test, type = "response")
glm.pred <- ifelse(glm.prob > 0.5, "High", "Low")
table(glm.pred, test$hiCrime)
mean(glm.pred == test$hiCrime)
```

This model gives a 89.5% test accruacy. Try a subset of data based on those that are statistically significant:

```{r}
glm.fit <- glm(hiCrime ~ nox + rad + ptratio, data = train, family = "binomial")
summary(glm.fit)
glm.prob <- predict(glm.fit, test, type = "response")
glm.pred <- ifelse(glm.prob > 0.5, "High", "Low")
table(glm.pred, test$hiCrime)
mean(glm.pred == test$hiCrime)
```

The more complex model performs better. Now try LDA for both sets of predictors:

```{r}
lda.fit <- lda(hiCrime ~ . - crim, data = train)
pred.lda <- predict(lda.fit, test)
table(pred.lda$class, test$hiCrime)
mean(pred.lda$class == test$hiCrime)

lda.fit <- lda(hiCrime ~ nox + rad + ptratio, data = train)
pred.lda <- predict(lda.fit, test)
table(pred.lda$class, test$hiCrime)
mean(pred.lda$class == test$hiCrime)
```

The logistic regression models outperform LDA based on test accuracy.

Now try KNN for different values of $K$:
```{r}
train_mat <- as.matrix(train[, -15])
test_mat <- as.matrix(test[, -15])
train_res <- ifelse(train$hiCrime == "High", 1, 0)

k2pred <- knn(train_mat, test_mat, train_res, k = 2)
k5pred <- knn(train_mat, test_mat, train_res, k = 5)
k10pred <- knn(train_mat, test_mat, train_res, k = 10)
k20pred <- knn(train_mat, test_mat, train_res, k = 20)

table(k2pred, ifelse(test$hiCrime == "High", 1, 0))
mean(k2pred == ifelse(test$hiCrime == "High", 1, 0))
table(k5pred, ifelse(test$hiCrime == "High", 1, 0))
mean(k5pred == ifelse(test$hiCrime == "High", 1, 0))
table(k10pred, ifelse(test$hiCrime == "High", 1, 0))
mean(k10pred == ifelse(test$hiCrime == "High", 1, 0))
table(k20pred, ifelse(test$hiCrime == "High", 1, 0))
mean(k20pred == ifelse(test$hiCrime == "High", 1, 0))
```
Of the values for $K$, 2 performs best, with a higher test accuracy than either logistic regression or LDA.
